---
title: "Understanding Alcohol Use in Canada"
author: "Dhanraj Patel"
date: '2021-12-17'
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
---
# setting up libraries needed
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(opendatatoronto)
library(MASS)
library(tidyverse)
library(janitor)
library(skimr)
library(visdat)
library(car)
library(leaps)
```



# loading and cleaning data
```{r}

data2018 <- read.csv("cchs-82M0013-E-2017-2018-Annual-component_F1.csv") 

finaldata2018 <- data2018 %>%
  select(-DHHDGLVG, -DHHDG611, -DHHDGLE5, -DOALC,-ALC_005, -ALC_010, -ALC_015, -ALC_020, -DOALW, -ALW_005, -ALW_010,-ALW_015, -ALW_020, -ALW_025, -ALW_030, -ALW_035, -ALW_040, -ALWDVDLY, -X)


finaldata2018 <-finaldata2018 %>% 
  filter(ALWDVWKY < 996)

finaldata2018 <-finaldata2018 %>% 
  filter(DHHGMS < 9)

finaldata2018 <-finaldata2018 %>% 
  filter(DHHDGHSZ < 9)

finaldata2018 <-finaldata2018 %>% 
  filter(ALCDVTTM < 9)

finaldata2018 <-finaldata2018 %>% 
  filter(DHH_OWN < 6)


data <- tibble::rowid_to_column(finaldata2018, "ID") # needed to separate dataset for validation



```

#seperating dataset into two halves for validation purposes
```{r}


set.seed(1)
train <- data[sample(1:nrow(data), 18152, replace=F), ]
test <- data[which(!(data$ID %in% train$ID)),] 

```

#removing ID variable column in datasets
```{r}

train<- train %>%
  dplyr::select(-ID)


test<- test %>%
  dplyr::select(-ID)

finaldata2018 <- train
```


#EDA of all potential variables in model
```{r}
par(mfrow=c(3,3))
hist(finaldata2018$DHH_SEX,main="Histogram of sex ", xlab="1:female 2: male", ylab="count", border="blue", col="green")
hist(finaldata2018$DHHGMS,main="Histogram of Martital Stutus",xlab="1:married 2:common Law 3:divorced: 4: single", ylab="count", border="blue", col="blue")
hist(finaldata2018$DHHDGHSZ,main="Histogram of Household size", xlab="Num of people in household", ylab="count", border="blue", col="orange")
hist(finaldata2018$DHHDGL12,main="Has kids in household",  xlab="0: none 1:one or more ", ylab="count", border="blue", col="red")
hist(finaldata2018$DHHGAGE,main="Histogram of age intervals", xlab="Age intervals", ylab="count", border="blue", col="yellow")
hist(finaldata2018$ALCDVTTM,main="Type of drinker", xlab="1:regular 2: non regular", ylab="count", border="blue", col="purple")
hist(finaldata2018$ALWDVLTR,main="Has experienced Long term effects", xlab="1: yes 2:no", ylab="count", border="blue", col="pink")
hist(finaldata2018$ALWDVSTR,main="Has experienced short term effects", xlab="1: yes 2:no", ylab="count", border="blue", col="grey")
hist(finaldata2018$DHH_OWN,main="Owns House", xlab="1:owns 2:rents", ylab="count", border="blue", col="black")



```

#numerical summaries for response variable

```{r}



avgstat <- finaldata2018 %>%
  dplyr::select(ALWDVWKY) %>%
  summarize(
  Count = n(),
  Min = min(ALWDVWKY),
  Q1 = quantile(ALWDVWKY, c(.25)),
  Median = median(ALWDVWKY),
  Q3 = quantile(ALWDVWKY, c(.75)),
  Max = max(ALWDVWKY), 
  IQR = quantile(ALWDVWKY, c(.75))- quantile(ALWDVWKY, c(.25)), 
  Mean = round(mean(ALWDVWKY), 2),
  Trimmean = round(mean(ALWDVWKY, trim = 0.1),2), 
  Var = round(var(ALWDVWKY)),
  SD = round(sd(ALWDVWKY),2), 
  Range = max(ALWDVWKY)-min(ALWDVWKY))

knitr::kable(avgstat, caption = "The trimmed mean is trimmed by 10 percent")


```



 


#seeing properties of model with all predictors
```{r}


fullmod <- lm(ALWDVWKY ~.,  data = finaldata2018)
summary(fullmod)

```

#identifying properties of model with all possible predictcor model
```{r}

# assumptions
pairs(finaldata2018[,-c(7)])

plot(finaldata2018$ALWDVWKY ~ fitted(fullmod), main="Y versus Y-hat", xlab="Y-hat", ylab="Y")
lines(lowess(finaldata2018$ALWDVWKY ~ fitted(fullmod)), lty=2)

r <- resid(fullmod)


plot(r ~ fitted(fullmod), main="title", xlab="x", ylab="res.")
plot(r ~ finaldata2018$DHH_SEX, main="title", xlab="x", ylab="res")
plot(r ~ finaldata2018$DHHGMS, main="title", xlab="x", ylab="res")
plot(r ~ finaldata2018$DHHDGHSZ, main="title", xlab="x", ylab="res")
plot(r ~ finaldata2018$DHHDGL12, main="title", xlab="x", ylab="res")
plot(r ~ finaldata2018$DHHGAGE, main="title", xlab="x", ylab="res")
plot(r ~ finaldata2018$ALCDVTTM, main="title", xlab="x", ylab="res")
plot(r ~ finaldata2018$ALWDVLTR, main="title", xlab="x", ylab="res")
plot(r ~ finaldata2018$ALWDVSTR, main="title", xlab="x", ylab="res")
plot(r ~ finaldata2018$DHH_OWN, main="title", xlab="x", ylab="res")

qqnorm(r)
qqline(r)

# VIF score
vif(fullmod)


```



#using best subset model sleection on full model
```{r}



best <- regsubsets(ALWDVWKY ~., data=finaldata2018, nbest=1)
summary(best)

```


#using backwards subset model sleection on full model

```{r}

stepAIC(lm(ALWDVWKY ~ ., data=finaldata2018),
scope=list(lower=lm(ALWDVWKY ~ 1, data=finaldata2018)),
direction = "backward", k=2)

finmod <- lm(ALWDVWKY ~ ALWDVSTR + ALWDVLTR + ALCDVTTM + DHHGAGE +DHH_SEX,  data = finaldata2018)
summary(finmod)

vif(finmod) # vif score of recommended model
```

#identifying properties of decided model
```{r}

#dataset with only decided predictors in final model
selecteddata <- finaldata2018 %>%
  dplyr::select(ALWDVWKY, ALWDVSTR, ALWDVLTR, ALCDVTTM, DHHGAGE, DHH_SEX)


#model with decided predictors
finmod <- lm(ALWDVWKY ~ ALWDVSTR + ALWDVLTR + ALCDVTTM + DHHGAGE +DHH_SEX,  data = selecteddata)


#checking assumptions of model with decided predictors
pairs(selecteddata[,-c(1)])



plot(selecteddata$ALWDVWKY ~ fitted(finmod), main="Y versus Y-hat", xlab="Y-hat", ylab="Y")
lines(lowess(selecteddata$ALWDVWKY ~ fitted(fullmod)), lty=2)

r <- resid(finmod)


plot(r ~ fitted(finmod), main="title", xlab="x", ylab="res.")
plot(r ~ selecteddata$DHH_SEX, main="title", xlab="x", ylab="res")
plot(r ~ selecteddata$DHHGAGE, main="title", xlab="x", ylab="res")
plot(r ~ selecteddata$ALCDVTTM, main="title", xlab="x", ylab="res")
plot(r ~ selecteddata$ALWDVLTR, main="title", xlab="x", ylab="res")
plot(r ~ selecteddata$ALWDVSTR, main="title", xlab="x", ylab="res")


qqnorm(r)
qqline(r)

vif(fullmod) #vif score

#boxcox of model
p <- powerTransform(cbind(selecteddata[,1] + 0.000000000001, selecteddata[,2]+ 0.000000000001, selecteddata[,3]+ 0.000000000001, selecteddata[,4]+ 0.000000000001, selecteddata[,5]+ 0.000000000001, selecteddata[,6])~ 1)

summary(p)

#applying boxcox transformations found earlier
selecteddata <- selecteddata %>%
  mutate( new_ALWDVWKY = (ALWDVWKY)**0.5)
selecteddata <- selecteddata %>%
  mutate( new_ALWDVSTR = (ALWDVSTR)**10)
selecteddata <- selecteddata %>%
  mutate( new_ALWDVLTR = (ALWDVLTR)**7)
selecteddata <- selecteddata %>%
  mutate( new_ALCDVTTM = (ALCDVTTM)**-5.5)
selecteddata <- selecteddata %>%
  mutate( new_DHH_SEX = (DHH_SEX)**0.5)

#dataset with only transformed variables
justinc <- selecteddata %>%
  dplyr::select(new_ALWDVWKY, new_ALWDVSTR, new_ALWDVLTR, new_ALCDVTTM, DHHGAGE, new_DHH_SEX)




#model with  transformed variables
finmod3 <- lm(new_ALWDVWKY ~ new_ALWDVSTR + new_ALWDVLTR + new_ALCDVTTM + DHHGAGE +new_DHH_SEX,  data = justinc)
summary(finmod3)

temp <- justinc #needed so pairwise can print more descreptive names isntead of variable names
colnames(temp) <- c("response", "Short term eff", "long term eff", "Type drinker", "AGE", "SEX")

pairs(temp[,-c(1)])



plot(justinc$new_ALWDVWKY ~ fitted(finmod3), main="Y versus Y-hat", xlab="Y-hat", ylab="Y")
lines(lowess(justinc$new_ALWDVWKY ~ fitted(finmod3)), lty=2)

r <- resid(finmod3)

par(mfrow=c(2,4))
plot(r ~ fitted(finmod3), main="Res plot fitted", xlab="fitted", ylab="res")
plot(r ~ justinc$new_DHH_SEX, main="res of SEX", xlab="SEX", ylab="res")
plot(r ~ justinc$DHHGAGE, main="res of AGE", xlab="AGE", ylab="res")
plot(r ~ justinc$new_ALCDVTTM, main="res of type of drinker", xlab="type", ylab="res")
plot(r ~ justinc$new_ALWDVLTR, main="res of Long term eff", xlab="Long term eff", ylab="res")
plot(r ~ justinc$new_ALWDVSTR, main="res of short terme eff", xlab="Short term eff", ylab="res")
qqnorm(r)
qqline(r)

vif(finmod3) # checking VIF
```



#outliers, leverage points and influential points
```{r}

#model
finmod3 <- lm(new_ALWDVWKY ~ new_ALWDVSTR + new_ALWDVLTR + new_ALCDVTTM + DHHGAGE +new_DHH_SEX,  data = justinc)


# leverage points
n <- length(justinc$new_ALWDVWKY)
p <- length(coef(finmod3))-1

h <- hatvalues(finmod3)
hcut <- 2*(p+1)/n

w1 <- which(h > hcut)




plot(justinc[,1]~justinc[,2], main="sqrt(Def) vs Temp", xlab="Temp", ylab="sqrt(Def)")
points(justinc[w1,1]~justinc[w1,2], col="red", pch=19)
plot(justinc[,1]~justinc[,3], main="sqrt(Def) vs Density", xlab="Density", ylab="sqrt(Def)")
points(justinc[w1,1]~justinc[w1,3], col="red", pch=19)
plot(justinc[,1]~justinc[,4], main="sqrt(Def) vs Rate", xlab="Rate", ylab="sqrt(Def)")
points(justinc[w1,1]~justinc[w1,4], col="red", pch=19)
plot(justinc[,1]~justinc[,5], main="sqrt(Def) vs Rate", xlab="Rate", ylab="sqrt(Def)")
points(justinc[w1,1]~justinc[w1,5], col="red", pch=19)
plot(justinc[,1]~justinc[,6], main="sqrt(Def) vs Rate", xlab="Rate", ylab="sqrt(Def)")
points(justinc[w1,1]~justinc[w1,6], col="red", pch=19)


#outliers
r <- rstandard(finmod3)

w2 <- which(r < -4 | r > 4)

par(mfrow=c(3,2))
plot(justinc[,1]~justinc[,2], main="sqrt(Def) vs Temp", xlab="Temp", ylab="sqrt(Def)")
points(justinc[w2,1]~justinc[w2,2], col="red", pch=19)
plot(justinc[,1]~justinc[,3], main="sqrt(Def) vs Density", xlab="Density", ylab="sqrt(Def)")
points(justinc[w2,1]~justinc[w2,3], col="red", pch=19)
plot(justinc[,1]~justinc[,4], main="sqrt(Def) vs Rate", xlab="Rate", ylab="sqrt(Def)")
points(justinc[w2,1]~justinc[w2,4], col="red", pch=19)
plot(justinc[,1]~justinc[,5], main="sqrt(Def) vs Rate", xlab="Rate", ylab="sqrt(Def)")
points(justinc[w2,1]~justinc[w2,5], col="red", pch=19)
plot(justinc[,1]~justinc[,6], main="sqrt(Def) vs Rate", xlab="Rate", ylab="sqrt(Def)")
points(justinc[w2,1]~justinc[w2,6], col="red", pch=19)


#influential points
Dcutoff <- qf(0.5, p+1, n-p-1)
D <- cooks.distance(finmod3)
which(D > Dcutoff)


DFFITScut <- 2*sqrt((p+1)/n)
dfs <- dffits(finmod3)
w3 <- which(abs(dfs) > DFFITScut)


DFBETAcut <- 2/sqrt(n)
dfb <- dfbetas(finmod3)
w4 <- which(abs(dfb[,1]) > DFBETAcut)
w5 <- which(abs(dfb[,2]) > DFBETAcut)
w6 <- which(abs(dfb[,3]) > DFBETAcut)
w7 <- which(abs(dfb[,4]) > DFBETAcut)
w8 <- which(abs(dfb[,5]) > DFBETAcut)
w9 <- which(abs(dfb[,6]) > DFBETAcut)

w <- unique(c(w3,w4,w5,w6,w7,w8,w9))

par(mfrow=c(3,2))
plot(justinc[,1]~justinc[,2], main="weekly drinks vs short term eff", xlab="short term eff", ylab="weekly drinks")
points(justinc[w,1]~justinc[w,2], col="red")
plot(justinc[,1]~justinc[,3], main="weekly drinks vs long term eff", xlab="long term eff", ylab="weekly drinks")
points(justinc[w,1]~justinc[w,3], col="red")
plot(justinc[,1]~justinc[,4], main="weekly drinks vs type drinker", xlab="type drinker", ylab="weekly drinks")
points(justinc[w,1]~justinc[w,4], col="red")
plot(justinc[,1]~justinc[,5], main="weekly drinks vs age", xlab="age", ylab="weekly drinks")
points(justinc[w,1]~justinc[w,5], col="red")
plot(justinc[,1]~justinc[,6], main="weekly drinks vs sex", xlab="sex", ylab="weekly drinks")
points(justinc[w,1]~justinc[w,6], col="red", pch=19)

```




#validation
```{r}

#variable means in final model for testing dataset
mean(test$DHH_SEX)
mean(test$DHHGAGE)
mean(test$ALWDVSTR)
mean(test$ALWDVLTR)
mean(test$ALCDVTTM)
mean(test$ALWDVWKY)

#variable means in final model for traing dataset
mean(train$DHH_SEX)
mean(train$DHHGAGE)
mean(train$ALWDVSTR)
mean(train$ALWDVLTR)
mean(train$ALCDVTTM)
mean(train$ALWDVWKY)


#dataset with only decided predictors in validating model
valdata <- test %>%
  dplyr::select(ALWDVWKY, ALWDVSTR, ALWDVLTR, ALCDVTTM, DHHGAGE, DHH_SEX)

#transforming variables in testing dataset as done to training
valdata <- valdata %>%
  mutate( new_ALWDVWKY = (ALWDVWKY)**0.5)
valdata <- valdata %>%
  mutate( new_ALWDVSTR = (ALWDVSTR)**10)
valdata <- valdata %>%
  mutate( new_ALWDVLTR = (ALWDVLTR)**7)
valdata <- valdata %>%
  mutate( new_ALCDVTTM = (ALCDVTTM)**-5.5)
valdata <- valdata %>%
  mutate( new_DHH_SEX = (DHH_SEX)**0.5)



# validation data with transformed variables
finval <-lm(new_ALWDVWKY ~ new_ALWDVSTR + new_ALWDVLTR + new_ALCDVTTM + DHHGAGE +new_DHH_SEX,  data = valdata)
summary(finval)


```


checking transformations and VIF of validation model
```{r}




#dataset with only decided predictors in validating model
transformedvaldata <- valdata %>%
  dplyr::select(new_ALWDVWKY, new_ALWDVSTR, new_ALWDVLTR, new_ALCDVTTM, DHHGAGE, new_DHH_SEX)

#validation model
finval <-lm(new_ALWDVWKY ~ new_ALWDVSTR + new_ALWDVLTR + new_ALCDVTTM + DHHGAGE +new_DHH_SEX,  data = valdata)



#checking assumptions of validation model
pairs(transformedvaldata[,-c(1)])

plot(transformedvaldata$new_ALWDVWKY ~ fitted(finval), main="Y versus Y-hat", xlab="Y-hat", ylab="Y")
lines(lowess(transformedvaldata$new_ALWDVWKY ~ fitted(finval)), lty=2)

r <- resid(finval)


plot(r ~ fitted(finval), main="title", xlab="x", ylab="res.")
plot(r ~ transformedvaldata$new_DHH_SEX, main="title", xlab="x", ylab="res")
plot(r ~ transformedvaldata$DHHGAGE, main="title", xlab="x", ylab="res")
plot(r ~ transformedvaldata$new_ALCDVTTM, main="title", xlab="x", ylab="res")
plot(r ~ transformedvaldata$new_ALWDVLTR, main="title", xlab="x", ylab="res")
plot(r ~ transformedvaldata$new_ALWDVSTR, main="title", xlab="x", ylab="res")


qqnorm(r)
qqline(r)

# VIF score
vif(finval)
```

#Checking influential points on validation
```{r}
# validation model and dataset
finval <-lm(new_ALWDVWKY ~ new_ALWDVSTR + new_ALWDVLTR + new_ALCDVTTM + DHHGAGE +new_DHH_SEX,  data = valdata)
transformedvaldata

# leverage points
n <- length(transformedvaldata$new_ALWDVWKY)
p <- length(coef(finval))-1

h <- hatvalues(finval)
hcut <- 2*(p+1)/n

w1 <- which(h > hcut)




plot(transformedvaldata[,1]~transformedvaldata[,2], main="sqrt(Def) vs Temp", xlab="Temp", ylab="sqrt(Def)")
points(transformedvaldata[w1,1]~transformedvaldata[w1,2], col="red", pch=19)
plot(transformedvaldata[,1]~transformedvaldata[,3], main="sqrt(Def) vs Density", xlab="Density", ylab="sqrt(Def)")
points(transformedvaldata[w1,1]~transformedvaldata[w1,3], col="red", pch=19)
plot(transformedvaldata[,1]~transformedvaldata[,4], main="sqrt(Def) vs Rate", xlab="Rate", ylab="sqrt(Def)")
points(transformedvaldata[w1,1]~transformedvaldata[w1,4], col="red", pch=19)
plot(transformedvaldata[,1]~transformedvaldata[,5], main="sqrt(Def) vs Rate", xlab="Rate", ylab="sqrt(Def)")
points(transformedvaldata[w1,1]~transformedvaldata[w1,5], col="red", pch=19)
plot(transformedvaldata[,1]~transformedvaldata[,6], main="sqrt(Def) vs Rate", xlab="Rate", ylab="sqrt(Def)")
points(transformedvaldata[w1,1]~transformedvaldata[w1,6], col="red", pch=19)


#outliers
r <- rstandard(finval)

w2 <- which(r < -4 | r > 4)


plot(transformedvaldata[,1]~transformedvaldata[,2], main="sqrt(Def) vs Temp", xlab="Temp", ylab="sqrt(Def)")
points(transformedvaldata[w2,1]~transformedvaldata[w2,2], col="red", pch=19)
plot(transformedvaldata[,1]~transformedvaldata[,3], main="sqrt(Def) vs Density", xlab="Density", ylab="sqrt(Def)")
points(transformedvaldata[w2,1]~transformedvaldata[w2,3], col="red", pch=19)
plot(transformedvaldata[,1]~transformedvaldata[,4], main="sqrt(Def) vs Rate", xlab="Rate", ylab="sqrt(Def)")
points(transformedvaldata[w2,1]~transformedvaldata[w2,4], col="red", pch=19)
plot(transformedvaldata[,1]~transformedvaldata[,5], main="sqrt(Def) vs Rate", xlab="Rate", ylab="sqrt(Def)")
points(transformedvaldata[w2,1]~transformedvaldata[w2,5], col="red", pch=19)
plot(transformedvaldata[,1]~transformedvaldata[,6], main="sqrt(Def) vs Rate", xlab="Rate", ylab="sqrt(Def)")
points(transformedvaldata[w2,1]~transformedvaldata[w2,6], col="red", pch=19)


#influential points
Dcutoff <- qf(0.5, p+1, n-p-1)
D <- cooks.distance(finval)
which(D > Dcutoff)


DFFITScut <- 2*sqrt((p+1)/n)
dfs <- dffits(finval)
w3 <- which(abs(dfs) > DFFITScut)


DFBETAcut <- 2/sqrt(n)
dfb <- dfbetas(finval)
w4 <- which(abs(dfb[,1]) > DFBETAcut)
w5 <- which(abs(dfb[,2]) > DFBETAcut)
w6 <- which(abs(dfb[,3]) > DFBETAcut)
w7 <- which(abs(dfb[,4]) > DFBETAcut)
w8 <- which(abs(dfb[,5]) > DFBETAcut)
w9 <- which(abs(dfb[,6]) > DFBETAcut)

w <- unique(c(w3,w4,w5,w6,w7,w8,w9))


plot(transformedvaldata[,1]~transformedvaldata[,2], main="weekly drinks vs short term eff", xlab="short term eff", ylab="weekly drinks")
points(transformedvaldata[w,1]~transformedvaldata[w,2], col="red")
plot(transformedvaldata[,1]~transformedvaldata[,3], main="weekly drinks vs long term eff", xlab="long term eff", ylab="weekly drinks")
points(transformedvaldata[w,1]~transformedvaldata[w,3], col="red")
plot(transformedvaldata[,1]~transformedvaldata[,4], main="weekly drinks vs type drinker", xlab="type drinker", ylab="weekly drinks")
points(transformedvaldata[w,1]~transformedvaldata[w,4], col="red")
plot(transformedvaldata[,1]~transformedvaldata[,5], main="weekly drinks vs age", xlab="age", ylab="weekly drinks")
points(transformedvaldata[w,1]~transformedvaldata[w,5], col="red")
plot(transformedvaldata[,1]~transformedvaldata[,6], main="weekly drinks vs sex", xlab="sex", ylab="weekly drinks")
points(transformedvaldata[w,1]~transformedvaldata[w,6], col="red", pch=19)
```

